From Reversible Programs to \\ Univalent Universes and Back

### 2021-07-09 Pi+/paper/discussion.tex
clean up univalence para in sec 8

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.

% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.

By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case).

@@

Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
\citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
$\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
subuniverses.

>>>>

%%%%%%%%%%%%%%
\paragraph{Computing with Univalence} In HoTT, univalence characterises the path type in the universe as equivalences of
types. The map $\term{idtoeqv} : A \id_{\UU} B \to A \eqv B$ can be easily constructed using path induction. The term
$\term{ua} : A \eqv B \to A \id_{\UU} B$, its computation rule $\term{ua_\beta} : (e : A \eqv B) \to
    \term{idtoeqv}(\term{ua}(e)) \id e$, and its extensionality rule $\term{ua_\eta} : (p : A \id_{\UU} B) \to p ==
    \term{ua}(\term{idtoeqv(p)})$ are generally added as postulates when formalising in Agda. Together, $\term{ua}$ and
$\term{ua}_\beta$ give the full univalence axiom $(A \eqv B) \eqv (A \id_{\UU} B)$.
% Let's think of the $\PiLang$ combinators as describing the inhabitants of the identity type of finite types.
By giving a computable presentation for a univalent subuniverse, we are able to describe its path space syntactically
via a complete equational axiomatisation of the equivalences between types in the subuniverse.
% By the property of being univalent, this subuniverse gives a model of the univalence axiom.
In the subuniverse of finite types, $\term{idtoeqv}$ corresponds to giving a denotation for a program (1-combinator),
which is easily done by induction.  The $\term{ua}$ map corresponds to synthesising a program from an equivalence
(which, in general, is of course undecidable~\cite{krogmeierDecidableSynthesisPrograms2020}). In case of reversible
boolean circuits, it is decidable, as we have shown, but still far from trivial, which matches the need to assert the
existence of $\term{ua}$ without giving a constructive argument. Then, the computation rule $\term{ua_\beta}$ expresses
the fact that program synthesis is sound, while $\term{ua_\eta}$ corresponds to the soundness of the equational theory
($\PiLang$ 2-combinators). Thus, our results suggests a new computational interpretation of the univalence principle, which
provides an intuition on why certain constructions are hard (or impossible in the general case).

@@

% Univalent Fibrations were introduced by~\citet*{kapulkinUnivalenceSimplicialSets2018},
% to build a model of Voevodsky's \emph{univalence} principle in simplicial sets.
% \citet{christensenCharacterizationUnivalentFibrations2015} studied characterisations of univalent fibrations using the
% $\BAut$ construction. Univalent typoids~\cite{petrakisUnivalentTypoids2019a} are a different presentation of univalent
% subuniverses.
